---
title: "Beat Yourself: New Haven Road Race 5K Times Compared to Self"
output: pdf_document
author: Michael Li and Davey Proctor
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r, echo=FALSE, results='hide', message=FALSE}
library(ggplot2)
library(dplyr)
df <- read.csv("NHRR_5K_Merge_Clean.csv", as.is=T)
divs <- df %>% distinct(Div)
divs[order(divs),]
# delete metadivs
df <- df[df$Div != "M01-19",]
df <- df[df$Div != "F01-19",]
```


```{r, echo=FALSE, results='hide', message=FALSE}
# Arrange by year and nettime
df %>% arrange(Year) %>% select(Year, Nettime)
# only runners who've run every year
superRunners <- df %>% group_by(Name) %>% filter(n() == 7) %>% select(Name, Div, Year, Nettime)
superRunners
```

# TODO: Group trends by age

## Group: Direct age effect plot
```{r, echo=FALSE, results='hide', message=FALSE}
df %>% ggplot(aes(x=Age, y=Nettime, color=Sex)) + geom_smooth()
# ggplot(df) + geom_smooth(data=df, aes(x=Age, y=Nettime, color=Sex), se=T)
```


# TODO: Relativize to runner
## Relativize: Explore
```{r, echo=FALSE, results='hide', message=FALSE}
set.seed(123)
names <- sample(superRunners$Name, 10)
superRunners[superRunners$Name %in% names,]
superRunners[superRunners$Name %in% names,] %>% ggplot(aes(x=Year, y=Nettime, col=Name)) + geom_line()
```

## Relativize: Stats per runner
```{r, echo=FALSE, results='hide', message=FALSE}
# lm on yearly difference for runner
andrea <- df[df$Name == "Andrea Benjamin",]
# andrea$Year <- andrea$Year - 2012 # not needed; affects the p-val of intercept but not slope of bestfit line.
m <- lm(Nettime ~ Year, data=andrea)
m$coefficients[["Year"]]
summary(m)
lm(Nettime ~ Year, data=andrea)$coefficients[["Year"]]
lm(andrea[["Nettime"]] ~ andrea[["Year"]])
# All together
#superRunners$Year <- superRunners$Year - min(superRunners$Year)
supRunChanges <- superRunners %>% group_by(Name) %>% summarise(effectDueToYear = lm(Nettime ~ Year)$coefficients[["Year"]])
supRunChanges
```

```{r, echo=FALSE, results='hide', message=FALSE}
summary(supRunChanges$effectDueToYear)
supRunChanges %>% ggplot(aes(x=effectDueToYear)) + geom_histogram(bins=50)
```

## Relativize: Null hypothesis: No effect due to age. Reject p <.001.
* Assume normal model
```{r, echo=FALSE, results='hide', message=FALSE}
z.test <- function(arr, x=0)
{
  n <- length(arr)
  xBar <- mean(arr)
  sdXBar <- sd(arr) / sqrt(n)
  pnorm(x, xBar, sdXBar)
}
nrow(supRunChanges)
mean(supRunChanges$effectDueToYear)
z.test(supRunChanges$effectDueToYear)
```

# TODO: Use full dataset.

```{r, echo=FALSE, results='hide', message=FALSE}
df %>% group_by(Name) %>% filter(n() == 2)
runChanges <- df %>% group_by(Name) %>% filter(n() > 2) %>% summarise(effectDueToYear = lm(Nettime ~ Year)$coefficients[["Year"]], Div=Div[[1]], Age=mean(Age), Sex=Sex[[1]], n=n(), meanNettime=mean(Nettime)) # >2 lest too heavy of tails; not remotely normal. also helps remove NAs in fits.
# runChanges[is.na(runChanges$effectDueToYear),]
# runChanges <- runChanges[!is.na(runChanges$effectDueToYear),]
```

## In general, people worsen year to year. p<1e-08.
```{r, echo=FALSE, results='hide', message=FALSE}
summary(runChanges$effectDueToYear)
runChanges[runChanges$effectDueToYear<20,] %>% ggplot(aes(x=effectDueToYear)) + geom_histogram(bins=50)
qqnorm(runChanges[runChanges$effectDueToYear<20,]$effectDueToYear)
nrow(runChanges)
mean(runChanges$effectDueToYear)
z.test(runChanges$effectDueToYear)
```

## Older people worsen fastest.
```{r, echo=FALSE, results='hide', message=FALSE}
women <- runChanges[runChanges$Sex=="F",]
ggplot(women, aes(x=Div, y=effectDueToYear)) + geom_boxplot()
ggplot(runChanges, aes(x=Age, y=effectDueToYear)) + geom_smooth()
```

## More consistent runners aren't more consistent; if anything they worsen faster than the others (inconclusive, though).
```{r, echo=FALSE, results='hide', message=FALSE}
# runChanges[runChanges$n > 7,] # This is odd
# df[df$Name == "Stephen Murphy",]
ggplot(runChanges[runChanges$n<=7 & runChanges$effectDueToYear<20,], aes(x=as.factor(n), y=effectDueToYear)) + geom_boxplot()
#ggplot(runChanges[1:500,], aes(x=as.double(n), y=effectDueToYear)) + geom_line()
t.test(runChanges[runChanges$n==3,]$effectDueToYear, runChanges[runChanges$n==7,]$effectDueToYear)
```

## Women aren't worsening any faster than men, and vice-versa.
```{r, echo=FALSE, results='hide', message=FALSE}
t.test(runChanges[runChanges$Sex=="F",]$effectDueToYear, runChanges[runChanges$Sex=="M",]$effectDueToYear)
```

## Worse runners worsen more year by year.
```{r, echo=FALSE, results='hide', message=FALSE}
ggplot(runChanges, aes(x=meanNettime, y=effectDueToYear)) + geom_smooth()
```


## Appendix
```{r ref.label=knitr::all_labels(), echo=T, eval=F}

```
